\documentclass[12pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx,psfrag,epsf}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{dcolumn}

\usepackage{color}
\usepackage{url}
% \usepackage[strings]{underscore}  % to handle _ in BibTeX URLs/DOIs; but messes up fig paths

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%To reset the counters at the start of each section and prefix the numbers by the section number
%(Based on: http://en.wikibooks.org/wiki/LaTeX/Tips_and_Tricks#Grouping_Figure.2FEquation_Numbering_by_Section)
\numberwithin{equation}{section}
\numberwithin{figure}{section}
\numberwithin{table}{section}

\DeclareMathOperator\erf{erf}

\def\sharedaffiliation{
\end{tabular} \begin{tabular}{c}}

\newcommand{\missingref}{\textcolor{red}{REF}}
\newcommand{\fix}[1]{\textcolor{red}{#1}}
\newcommand{\dd}{\textrm{d}}
% From the C++ FAQ:
\newcommand{\Cpp}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}

% Do NOT hyphenate all-caps words:
%\uchyph=0
\hyphenation{CUDA CUDAHM}


\newcolumntype{d}[1]{D{.}{.}{#1}} \newcolumntype{t}[1]{D{-}{\times}{#1}}

\newcommand{\assm}{\,|\,}
\newcommand{\mlike}{\ell}
\newcommand{\ppar}{\theta}  % pop'n parameters
\newcommand{\obsv}{\mathcal{O}}  % observables
\newcommand{\psivec}{\boldsymbol{\psi}}
\newcommand{\Dvec}{\boldsymbol{D}}
\newcommand{\lfunc}{\phi}  % or \Phi or \varphi or...
\newcommand{\clfunc}{\Phi}
\newcommand{\lpdf}{f}
\newcommand{\lcdf}{F}
\newcommand{\rhopar}{\boldsymbol{\zeta}}
\newcommand{\dtxns}{\mathcal{D}}
\newcommand{\Fth}{F_{\rm th}}
\newcommand{\Fhat}{\hat{F}}
\newcommand{\Ffid}{F_{\rm fid}}
\newcommand{\rmax}{r_{\rm max}}
\newcommand{\effic}{\eta}
\newcommand{\aeffic}{\overline{\eta}}
\newcommand{\like}{\mathcal{L}}
\newcommand{\epdf}{\mu}
\newcommand{\Lsol}{L_\odot}

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% ONLY NEEDED FOR DRAFTS (delete or comment for submission):

% Macro for an "editorial note"; uncomment the 2nd definition to
% produce a version without notes (put it in main doc if desired).
\newcommand\enote[1]{{$\bullet\bullet\bullet$}{\sl [#1]}{$\bullet\bullet\bullet$}}
%\renewcommand\enote[1]{\relax}

% Marginal note (just a redef of \marginpar or \marginparodd).
% Uncomment the 2nd definition to produce a version without notes (put it 
% in main doc if desired).
% Could use \footnotesize for a bigger font.
% The \hspace enables hyphenation of the 1st word.
\setlength{\marginparwidth}{.85in}
\newcommand\mnote[1]{\-\marginpar[\raggedleft\scriptsize\hspace*{0pt}#1]%
{\raggedright\scriptsize\hspace*{0pt}#1}}
%\renewcommand\mnote[1]{\relax}

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

%===============================================================================
\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\if0\blind
{
  %\title{\bf CUDAHM:  GPU-Accelerated Bayesian Inference for Simple Hierarchical Models}
  %\title{\bf CUDAHM:  GPU-Accelerated Hierarchical Bayesian Modeling of Cosmic Populations}
  \title{\bf CUDAHM:  GPU-Accelerated Hierarchical Bayesian Inference
  with Application to Modeling Cosmic Populations}
  \author{J\'anos M. Szalai-Gindl\\
    %\thanks{
    %\textit{This work was supported by the Hungarian Scientific Research Fund via grant OTKA~NN~114560.  Budav\'ari, Kelly, and Loredo gratefully acknowledge the NSF-funded Statistical and Applied Mathematical Sciences Institute (SAMSI) for support for visits to SAMSI, where this project originated. Loredo's effort was additionally supported by NSF grant AST-1312903.}}\hspace{.2cm}\\
    Department of Information Systems,\\
    E\"otv\"os Lor\'and University\\
    and \\
    Tam\'as Budav\'ari \\
    Department of Applied Mathematics \& Statistics,\\
    The Johns Hopkins University\\
    and \\
    Brandon C. Kelly \\
    Department of Physics,\\
    Broida Hall, University of California\\
    and \\
    Thomas J. Loredo \\
    Cornell Center for Astrophysics \& Planetary Science,\\
    Cornell University \\%    and \\
%    Istv\'an Csabai\\
%    Department of Physics of Complex Systems,\\
%    E\"otv\"os Lor\'and University\\
%    and \\
%    L\'aszl\'o Dobos\\
%    Department of Physics of Complex Systems,\\
%    E\"otv\"os Lor\'and University
}
    
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf CUDAHM:  GPU-Accelerated Hierarchical Bayesian Inference}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
We describe CUDAHM, a \Cpp\ framework for hierarchical Bayesian inference with simple (typically single-plate) parametric graphical models that uses graphics processing units (GPUs) to accelerate computations, enabling deployment on very large datasets.
CUDAHM exploits conditional independence between instances of a plate, which enables massively parallel exploration of the replication parameter space using the single instruction, multiple data (SIMD) architecture of GPUs.
It provides support for constructing Metropolis-within-Gibbs samplers that iterate between GPU-acclerated robust adaptive Metropolis (RAM) sampling of plate-level parameters conditional on upper-level parameter values, and Metropolis-Hastings sampling of upper-level parameters on the host central processing unit (CPU) conditional on the GPU results.
The GPU computations are implemented using the Compute Unified Device Architecture (CUDA) parallel computing platform.
CUDAHM is motivated by measurement error problems in astronomy, where density estimation and linear and nonlinear regression problems must be addressed for populations of thousands to millions of objects whose features are measured with possibly complex uncertainties.
We demonstrate accurate GPU-accelerated parametric conditional density deconvolution for simulated populations of $10^6$ objects in about two hours using a single NVIDIA Tesla K40c GPU.
Supplementary material provides details about the demonstration problem.
\end{abstract}

\noindent%
{\it Keywords:}
Hierarchical Bayesian models, Metropolis-within-Gibbs sampling, parallel computing, astrostatistics, graphical processing units (GPUs)
\vfill

\newpage \spacingset{1.45} % DON'T change the spacing!

\input{Intro}

\input{Design}

%===============================================================================
%\section{Example: Dust properties in a star-forming region}
%\label{sec:dust}
%
%\enote{Add material here for Brandon's dust SED example if desired; we should probably try to save this for the A\&C version.}
%===============================================================================

% These perhaps should be merged:
\input{LumFuncA}
%\input{TEMP}
\input{LumFuncB}

\input{Discussion}


%===============================================================================
\begin{center}
{\large\bf ACKNOWLEDGMENTS}
\end{center}

This material was based upon work partially supported by the National Science Foundation under Grant DMS-1127914 to the Statistical and Applied Mathematical Sciences Institute (SAMSI). 
The CUDAHM project began in a working group affiliated with the 2012--2013 SAMSI Program on Statistical and Computational Methodology for Massive Datasets.
This paper was completed as part of the 2016--2017 SAMSI Program on Statistical, Mathematical and Computational Methods for Astronomy.
Budav\'ari, Kelly, and Loredo are grateful to SAMSI for support of this work.
Loredo's work on this project was also supported by NSF grant AST-1312903.
Szalai-Gindl's work at Johns Hopkins University was supported by the Hungarian Scientific Research Fund via grant OTKA~NN~114560, where he was a visiting graduate student.
Any opinions, findings, and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation.


%===============================================================================
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

A supplementary document contains two appendices providing further details about the luminosity function application described in Sections \ref{sec:lum_func} and \ref{sec:lum_func_sim}.
\begin{description}
\item[Appendix A:] Thinned latent marked point process likelihood function
\item[Appendix B:] Exponential-cutoff break-by-one power law luminosity function
\end{description}


%===============================================================================
\bibliographystyle{agsm}

\bibliography{CUDAHM}

\end{document}
